\documentclass[a4paper,12pt]{letter}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nopageno}
\usepackage{amsmath}
%\usepackage{enumitem}
% \usepackage{mathtools}
%\usepackage{eqnarray,amsmath}
\usepackage[a4paper, total={7.5in, 11.25in}]{geometry}
%\usepackage[margin=2.5cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{amsmath, amssymb}
\usepackage{setspace}
\setstretch{1.2}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}


\begin{document}

\begin{center}
 (KTXFI1EBNF) Physics I.  Homework No.~1\\ \smallskip
March~8,~2026
 \end{center}

\begin{enumerate}

\item A projectile is launched at an angle of $30^\circ$ with an initial velocity of $30\,\text{m/s}$. 
Calculate:
\begin{enumerate}
    \item[(a)] The time to reach the peak of the trajectory.
    \item[(b)] The maximum height.
    \item[(c)] The horizontal range (impact distance).
    \item[(d)] The total time of flight.
\end{enumerate}

\item At what distance from the Mars does an areosynchronous satellite orbit in the plane of the Equator if it remains constantly above the same point on the Mars's surface?

\item A $2\,\text{kg}$ object slides down from the top of a $3\text{-meter-high}$ inclined plane that makes an angle of $30^\circ$ with the horizontal. What is the velocity of the object when it reaches the bottom of the incline, starting from rest at the top, if:
\begin{enumerate}
    \item[(a)] there is no friction between the object and the incline?
    \item[(b)] the coefficient of friction is $0.05$?
\end{enumerate}

\item A string is wound around a solid cylinder with radius $R = 0.5\,\text{m}$ and mass $M = 30\,\text{kg}$. Two objects with a combined mass of $15\,\text{kg}$ are hanging from the ends of the string. Upon release, the cylinder accelerates with an angular acceleration of $\beta = 4\,\text{rad/s}^2$. The string does not slip.
\begin{itemize}
    \item[(a)] Find the individual masses ($m_1, m_2$) of the two objects.
    \item[(b)] Calculate the total kinetic energy of the system at $t = 4\,\text{s}$.
\end{itemize}

\item Estimate the average depth of the Pacific Ocean if a tsunami takes 5 hours to travel from Hawaii to Alaska. (Note: The propagation speed of shallow-water waves is $v=\sqrt{gh}$. The distance is $4000$\,km.)


\end{enumerate}

  
\bigskip

\rightline{Dr.~Gabor~FACSKO}
\rightline{\textit{facsko.gabor@uni-obuda.hu}}

\vfill

\end{document}
